### 2020-09-10 Manuscript/manuscript.tex
progress on adding evaporite paleolatitudes to analysis and manuscript

Sedimentary rocks whose deposition is associated with certain climatic conditions have the potential to provide insight into paleolatitude. 
Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation. Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003a}. Evaporite deposits are particularly compelling as a paleolatitude constraint given that their deposition is associated arid regions resulting from large scale Hadley cell downwelling \citep{Evans2006a}. While moisture in the subtropics can change along with Earth's climate the overall pattern of $\sim$15-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}. Using compiled data that paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga that large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones. This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale Hadley cell circulation.

Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
\begin{itemize}
%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}. The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited 
\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict preserved gypsum and anhydrite within secondary silica \citep{White1984a}. The correlative Prichard Formation is 
intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}. Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}. Further up in the Belt Supergroup , within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}. These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and likely similar in age to the 1454 $\pm$ 9 Ma tuff within Helena Formation \citep{Evans2000c}. 
\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated) and shale with halite casts \citep{Kah2001a}.  These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}  While gypsum beds can reach thicknesses of 2.5 meters \citep{Kah2001a}, evaporites in the Iqqittuq Formation are interpreted to be volumetrically minor compared to units such as the Tonian Ten Stone Formation \citep{Turner2009a}.
\item the Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} . These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca. 1005 Ma detrital zircons \citep{Turner2016a}. The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca. 820 Ma \citep{Turner2016a}. These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
\item the Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}. The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum bearing siltstone\citep{Jefferson1989a, Jones2010a}. The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation.
\end{itemize}

>>>>

Sedimentary rocks whose deposition is associated with specific climatic conditions have the potential to provide insight into paleolatitude. Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation. Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003a}. Evaporite deposits are particularly compelling as paleolatitude constraints given that their deposition is interpreted to be associated with arid regions resulting from large-scale Hadley cell downwelling \citep{Evans2006a}. While moisture in the subtropics can change along with Earth's climate, the overall pattern of $\sim$10-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}. Using a compilation of paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones. This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale convection circulation cells.

There is high evaporation in the subtropics and tropics. Within the tropical rain belt (0 to $\sim$10\textdegree$\;$latitude) these high evaporation rates are typically overwhelmed by precipitation such that global zonal mean precipitation exceeds evaporation within $\sim$10\textdegree$\;$ of the equator (within $\sim$8\textdegree$\;$ of the equator over land) with evaporation exceeding precipitation from those latitudes towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree$\;$ \citep{Park2020b}. However, continental interiors near the equator can also be arid due to regional precipitation patterns leading to the precipitation of evaporites. For example, Lake Magadi in Kenya at a latitude of 1.9\textdegree$\;$S is a saline lake where thick bedded evaporites have accumulated \citep{Eugster1980a}. Caution is therefore needed when interpreting paleolatitude from evaporites in terrestrial and intracratonic settings given that they could occur both in tropical and subtropical latitudes.

Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
\begin{itemize}
%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}. The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited 
\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preservedwithin secondary silica \citep{White1984a}. The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}. Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}. Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}. There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.  These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 within the Helena Formation \citep{Evans2000c}. 
\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated with beds that reach a thickness of 2.5 meters) and shale with halite casts \citep{Kah2001a}.  These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} . These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca. 1005 Ma detrital zircons \citep{Turner2016a}. The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca. 820 Ma \citep{Turner2016a}. These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
\item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}. The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}. The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation.
\end{itemize}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_evaporite_figure.pdf}
\caption{\small{\textbf{Left panel: The paleolatitude of evaporite deposits following the amalgamation of Laurentia as reconstructed by the simple Laurentia model shown in Fig. \ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic. Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}. The evaporite lines extend from the maximum to minimum age constraints while the points are at the preferred depositional age.  Right panel: The present-day location of evaporites reconstructed to the left. 1: Altyn Formation (Belt Supergroup), 2: Wallace/Helena Formations (Belt Supergroup), 3: Iqqittuq Formation (Bylot Supergroup), 4: Ten Stone Formation (Mackenzie Mountains Supergroup), 5: Minto Inlet Formation (Shaler Supergroup), 6: Redstone River Formation (Mackenzie Mountains Supergroup), 7: Kilian Formation (Shaler Supergroup), 8: Carboniferous Canadian Maritime, 9: Carboniferous Sverdrup, 10: Permian Midcontinental USA.}}}
\label{fig:Laurentia_evaporites}
\end{figure}

In Figure \ref{fig:Laurentia_evaporites}, the paleolatitude of these evaporite deposits are reconstructed using the
